A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-29 paper.tex

There are two distinct annotation schemes possible, and the
distinction is subtle and important.
The first scheme is to use the annotations to describe
the complete set of genomic intervals inherited by the descendant node from the
ancestor node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to describe
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).

@@

ARGs are inherently a retrospective structure defined in terms
of a set of sample nodes (usually the leaves), and,
as we discuss in section XXX, storing sample-resolved ancestry
intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
useful than storing the local parent-child intervals of inheritance
(Fig~\ref{fig-ancestry-resolution}B).
Indeed, storing local parent-child
inheritance intervals for edges is really only of interest because of
its direct correspondence with the classical approach of associating
breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
The eARG encoding is simply a more limited form, in which the only types of direct
inheritance we can have are the entire genome, or the intervals
at either side of a recombination breakpoint.

@@

Given the local inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the ancestral material for a node is the
union of the ancestral intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their local inheritance intervals
and the just-computed ancestral intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries ancestral material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries ancestral
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. This process is directly analogous to Hudson's
simulation algorithm
(Appendix XXX).

>>>>

There are two distinct annotation schemes possible, and the distinction is subtle
and important. The first scheme is to use the annotations to describe the complete
set of genomic intervals directly inherited by the descendant node from the ancestor
node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).

@@

ARGs are inherently a retrospective structure defined in terms
of a set of sample nodes (usually the leaves), and,
as we discuss in section XXX, storing sample-resolved ancestry
intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
useful than storing the parent-child intervals recording direct inheritance
(Fig~\ref{fig-ancestry-resolution}B).
Indeed, storing direct inheritance intervals for edges is really only of interest because of
its correspondence with the classical approach of associating
breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
The eARG encoding is simply a more limited form, in which the only types of direct
inheritance we can have are the entire genome, or the intervals
at either side of a recombination breakpoint.

@@

Given the direct-inheritance edge annotations
(Fig~\ref{fig-ancestry-resolution}B),
 we obtain the sample-resolved edge annotation
(Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
Proceeding in topological order (such that an ancestor node is visited
after all of its descendants) the sample-resolved material for a node is the
union of the sample-resolved intervals on its inbound edges. We then annotate
the outbound edges with the intersection of their direct-inheritance intervals
and the just-computed sample-resolved intervals for the node.
For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
carries sample-resolved material on the intervals $(0, 2]$ and $(4, 7]$. The edge
joining \textsf{i} to its ancestor \textsf{k} then carries
material on $(0, 2]$ since although \textsf{i} directly inherits interval
$(0, 3]$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $(0, 2]$ is
ancestral to the sample. This process is directly analogous to Hudson's
simulation algorithm
(Appendix XXX).
